---
title: "Data Transformation Assignment"
output: github_document
---
```{r}
library(nycflights13)
library(tidyverse)
```
To access native functions that tidyverse overwrites use stats::filter() and stats::lag()
view flights opens the whole dataset in RStudio viewer
int - integers, dbl - double(realnumber), chr - character vectors, or strings, dttm (date+time)
lgl - logical (TRUE or FALSE)
date - Should be self evident

Variables are known by their columns
Observations by rows
Values are contained in each row column cell

You pick observations by their values with filter()
reorer rows by arrange()
pick variables by select()
Create new variables with mutate()
summarize with summarise()
group_by() - changes scope

First argument is data frame, everything else describes what to do
Save results by assigning to variable
(dec25 <- filter(flights, month == 12, day == 25))
Floating point numbers don't work well with == , use near() instead
near(sqrt(2)^2, 2)

& and
| or
! not
xor(x,y)

NA is a missing value
is.na(x) looks for missing values

#5.2.4 Exercises
```{r}
filter(flights, arr_delay > 2)
filter(flights, dest == 'IAH' | dest == 'HOU')
filter(flights, carrier == 'AA' | carrier == 'UA' | carrier == 'DL' )
filter(flights, (month == 7) | (month == 8) | (month == 9))
filter(flights, (arr_delay > 120) & (dep_delay <= 0))
filter(flights, (dep_delay >= 60) & (arr_delay <= 30))
filter(flights, (dep_time >= 2400) | (dep_time <= 600))

#This bit of code doesn't seem to work, it is selecting on 7
#Doesn't look like it works any way applied to month
#filter(flights, month %in% 7:9)
filter(flights, between(month, 7, 9))
#filter(flights, month %in% c(7, 8, 9))
filter(flights, between(hour, 0, 6))

filter(flights, is.na(dep_time))
```
Na ^ 0 equals 1. or true is always true. False and NA are not missing for a similar reason, the boolean expression
returns a T or F value, no matter what NA's value is. It appears from google the answer is way more complicated once
exceptions are included. Why is na^0 = 1 when na * 0 is NA...

MORE EXERCISES
```{r}
#arrange(flights, desc(is.na(flights?)))
arrange(flights, desc(dep_delay))
arrange(flights, dep_delay)
arrange(flights, air_time)
arrange(flights, desc(air_time))

```
Select all columns from year to day select(flights, year:day)
select all columns except those from year to day select(flights, -(year:day))

Helper functions for select include
starts_with("abc")
ends_with("xyz")
contains("ijk")
matches("(.)\\1")
num_range("x", 1:3)   Matches x1, x2, x3
everything()     selects all the variables remaining, useful to move a few things around

Select can remane variables, but drops all unselected ones
rename() keeps all other variables and renames one

EXERCISES
```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)
select(flights, dep_time, dep_time)
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
#select(flights, vars)
#select(flights, one_of(vars))
select(flights, contains("TIME"))
select(flights, contains("TIME", ignore.case = F))

```
It seems to drop any additional selections
One_of() returns an error "Error: All arguments must be character vectors, not list"
It is suprising...why is it not case sensitive?


mutate() makes new columns that are functions of other existing columns.

```{r}
flights_sml <- select(flights, year:day, ends_with("delay"), distance, air_time)
mutate(flights_sml, gain = dep_delay - arr_delay, speed = distance /air_time * 60)
```

transmute() only keeps the new values


```{r}
transmute(flights, gain = dep_delay - arr_delay, hours = air_time / 60, gain_per_hour = gain /hour)
```

cumulative and rolling aggregates
cumsun()
cumprod()
cummin()
cunmax()
cummean()
#These might be useful for networking class stats ^
min_rank()
row_number()
dense_rank()
percent_rank()
cume_dist()

# 5.5.2

```{r}
transmute(flights, hour * 60 + minute)
transmute(flights, arr_time-dep_time)
```




